Design and Implementation of Probabilistic Programming Language Anglican

### 2017-02-07 paper/IFL_2016_reviews_RESPONSE.txt

1) Add a clear and concise statement of the contributions up front.

Added a bullet list of contributions to the end of introduction.

2) The treatment of related work must be significantly improved.

Added section 2. Related work.

3) A clear and concise overview of the constructs of the language has
to be added, along with a good characterisation of at least the
semantics of the core constructs "sample" and "observe", as formal
as possible.

Rewrote section 5. Language, with the syntax of Clojure forms
supported, core library support, and section 5.3 on
probabilistic forms, with details on their syntax and semantics, 
related to section 4, the definition of probabilistic program.

4) Clarify why the language constructs justify building a new language
instead of making a library.

Added a clarification at the end of subsection 5.3.

5) Clearly bring out the central ideas and techniques of implementing
the core constructs "sample" and "observe" as independently as
possible of the specifics of Clojure so as to facilitate using
these in other (embedded) DSL implementations or, if feasible,
as libraries.

Rewrote subsection 7.2 'Probabilistic Forms' of section 7.
Macro-based compilation to provide more details and techniques
and separate Clojure dependent and language independent issues.

6) Reduce the stress (detail) on the *general* CPS based compilation techniques
so as to allow an improved coverage of the core primitives or other 
aspects that need further explanation.

Removed description of general CPS based compilation techniques.

7) Make it clear that section addressing of checkpoints refers to a known
addressing scheme, and reduce amount of detail if necessary to save
space, referring to earlier work instead.

Upon discussing this point with the reviewers we discovered that
there was misunderstanding, and addressing of checkpoints is
indeed a contribution of the paper:

> The reviewers have re-read this part, and it is now clear there was
> a misunderstanding on their part.
> 
> Thus, ignore point 7 as it stands, and instead just make it clear
> both up-front and where the addressing scheme is discussed that this
> is novel and one of the central contributions of the paper.
> 
> Additionally, in general, try to bring out the central contributions
> and ideas in a way that does not get obscured by too many details
> specific to your setting. I believe many people who seek to implement
> similar languages can learn a lot from your work, and it would be
> great if that is made as easy as possible!

Added addressing of checkpoints to the list of contribution in
the beginning of the paper. In subsection 8.3, Addressing of
checkpoint, changed the language 'uses a different scheme' to
'introduces a new scheme'.

8) Bring the convincing example at page 9 significantly forward.

Case study is now Section 6, right after language definition and
before discussing any implementation issues.

9) Substantiate performance claims by at least some concrete results
from benchmarking.

Added section 10 Performance, with performance comparison
between Anglican and Clojure, as well as between Anglican and
other probabilistic programming environment, with numbers.

10) If more space needs to be saved, remove the two paragraphs starting
"There are several ways to build a programming language..."

Removed the aforementioned paragraphs as well as other content
which is either too specific or trivial for the audience of IFL.

### 2017-02-08 paper/paper.tex
reformatted a paragraph, fixed a typo
more performance evaluation, shorter case study

\usepackage{listings}
\usepackage{lmodern}
\ttfamily
\DeclareFontShape{OT1}{lmtt}{m}{it}
     {<->sub*lmtt/m/sl}{}
\lstdefinestyle{default}{language=Anglican, basicstyle=\linespread{0.92}\ttfamily, columns=flexible, showstringspaces=false}

@@

For inference, we chose to use  Lightweight
Metropolis-Hastings, perhaps somewhat arbitrarily. A strength of
probabilistic programming is that models are separated from
inference. To switch to a different inference algorithm we just
need to pass a different value to \texttt{doquery}. For example,
we may decide to use Black-Box Variational Bayes (BBVB), which
may not work equally well for all probabilistic programs, but is
much faster to converge.

\begin{lstlisting}[style=default]
(def samples (doquery :bbvb deli nil))
\end{lstlisting}

We can still use samples and summary statistics with BBVB to
approximate the posterior distribution. However, variational
inference approximates the posterior by known distributions,
and we can directly retrieve the distribution parameters.

\begin{lstlisting}[style=default]
(clojure.pprint/pprint
  (anglican.bbvb/get-variational 
    (nth samples N)))
\end{lstlisting}

\begin{lstlisting}[style=default]
{S28209
 {(0 anglican.runtime.normal-distribution)
  {:mean 10.99753360180663,
   :sd 0.7290976433082352}},
 S28217
 {(0 anglican.runtime.normal-distribution)
  {:mean 12.668050292254202,
   :sd 0.9446695174790353}},
 S28215
 {(0 anglican.runtime.normal-distribution)
  {:mean 9.104132559955836,
   :sd 0.9479290526821788}},
 S28219
 {(0 anglican.runtime.flip-distribution)
  {:p 0.11671977016875924,
   :dist {:min 0.0, :max 1.0}}}}
\end{lstlisting}

We can guess that the variational distributions correspond to the
prior distributions in the \texttt{sample} forms, in the order
of appearance. However, it would help if we could use more
informative labels instead of automatically generated symbols
\texttt{S28209}, \texttt{S28217}, etc. Here the option to
specify identifiers for probabilistic forms explicitly comes
handy. If we modify the \texttt{sample} forms to use explicit identifiers
(only the forms are shown for brevity), the output becomes much
easier to analyse.

\begin{lstlisting}[style=default]
(defm same-customer 
      ...
      (sample :arrival-time-same
              time-to-arrive-prior)]
      ...)

(defm different-customers
      ... 
      (sample :arrival-time-first
              time-to-arrive-prior)
      ...
      (sample :arrival-time-second
              time-to-arrive-prior)]
      ...)

(defquery deli ...
      ...
      (sample :same-or-different
              (flip p-same))
      ...)
\end{lstlisting}

The output becomes much easier to interpret and analyse
programmatically.

\begin{lstlisting}[style=default]
{:arrivate-time-same
 {(0 anglican.runtime.normal-distribution)
  {:mean 10.99753360180663,
   :sd 0.7290976433082352}},
 :arrival-time-first
 {(0 anglican.runtime.normal-distribution)
  {:mean 12.668050292254202,
   :sd 0.9446695174790353}},
 :arrival-time-second
 {(0 anglican.runtime.normal-distribution)
  {:mean 9.104132559955836,
   :sd 0.9479290526821788}},
 :same-or-different
 {(0 anglican.runtime.flip-distribution)
  {:p 0.11671977016875924,
   :dist {:min 0.0, :max 1.0}}}}
\end{lstlisting}

This completes the case study, where we showed a probabilistic
programming solution of a problem, implemented and analysed in
Anglican and Clojure, using two different inference algorithms.

@@

We run this program for n equal to 10, 15, 20, and 25. Clojure
is consistently only twice as fast as Anglican
(Table~\ref{tbl:hanoi}); considering the amount of overhead
Anglican introduces, this is a good result.
\begin{table}
    \centering
    \begin{tabular}{r|r|r|r|r}
        {\bf n}  & {\bf 10} & {\bf 15} & {\bf 20} &{\bf 25} \\ \hline
        Clojure &  0.32 & 1.45 & 38.1  &  1245.9 \\ \hline
        Anglican & 0.63 & 3.00 & 75.9 & 2381.1 
    \end{tabular}
    \caption{Towers of Hanoi. Average running times, in
    seconds, of Clojure and Anglican, averaged over 100 runs.
    Clojure is only twice as fast as Anglican.}
    \label{tbl:hanoi}
\end{table}

The other comparison must be made between Anglican and other
probabilistic programming environment. \cite[pp. 32 -- 33]{P16} 
provides a comparison of Anglican to an older, interpreted
version of Anglican~\cite{WVM14}, and Probabilistic
Scheme~\cite{PW14} on a rather complicated inference task of
program synthesis. According to this comparison, Anglican is at
least 10 times faster than the older interpreted version, and
almost as fast as Probabilistic Scheme which uses inference
engine implemented in C. In addition, \cite[p. 171]{SGG15}
compares Anglican to a DSL for probabilistic programming in
Haskell and to Probabilistic C~\cite{PW14} on two simple
probabilistic programs. Anglican shows similar performance
to the Haskell implementation of Sequential Monte
Carlo~\cite{DFG+01} and scales better with the number of
particles. Given the simplicity of the programs used in the
evaluation, this comparison is a confirmation that the flexible
checkpoint and trampoline based interaction between an Anglican
program and the inference engine does not introduce any noticeable
overhead compared to more rigid designs.

>>>>

\usepackage{listings}
\usepackage{lmodern}
\ttfamily
\DeclareFontShape{OT1}{lmtt}{m}{it}
     {<->sub*lmtt/m/sl}{}
\lstdefinestyle{default}{language=Anglican,basicstyle=\linespread{1.0}\ttfamily, columns=flexible, showstringspaces=false}

@@

\iftoggle{full}{%
\input{case-study-bbvb}
}{}%

This completes the case study, where we showed a probabilistic
programming solution of a problem, implemented and analysed in
Anglican and Clojure\iftoggle{full}{, using two different
inference algorithms}{}.

@@

We run this program for n equal to 10, 15, 20, and 25. Clojure
is consistently only twice as fast as Anglican
(Table~\ref{tbl:hanoi}); considering the amount of overhead
Anglican introduces, this is a good result.
\begin{table}
    \centering
    \begin{tabular}{r|r|r|r|r}
        {\bf n}  & {\bf 10} & {\bf 15} & {\bf 20} &{\bf 25} \\ \hline
        Clojure &  0.32 & 1.45 & 38.1  &  1245. \\ \hline
        Anglican & 0.63 & 3.00 & 75.9 & 2381. 
    \end{tabular}
    \caption{Towers of Hanoi. Running times, in
    seconds, of Clojure and Anglican, averaged over 100 runs.
    Clojure is only twice as fast as Anglican.}
    \label{tbl:hanoi}
\end{table}

The other comparison must be made between Anglican and another
probabilistic programming environment. WebPPL~\cite{GS15} is
similar to Anglican in expressivity.  WebPPL runs on top of
\href{http://v8project.blogspot.com/}{V8}, a high performance
JavaScript engine featuring a native compiler.
Table~\ref{tbl:webppl} shows results for two models from WebPPL
distribution, \textit{LDA} and
\textit{Logistic regression}. For
comparison, we re-implemented the models in Anglican, as close
as possible to the WebPPL implementations, and run the same
inference algorithm (called MCMC in WebPPL, LMH in Anglican)
with the same number of iterations. Anglican outperforms WebPPL
on both models. This can be attributed both to a better
transformed representation of probabilistic programs and to more
efficient architecture of the inference engine.
\begin{table}
    \centering
    \begin{tabular}{r|c|c}
        {\bf model}  & {\bf LDA} & {\bf Logistic regression} \\ \hline
        Anglican & 22. & 4.8 \\ \hline
        WebPPL & 31. & 9.4
    \end{tabular}

    \caption{Anglican vs. WebPPL. Running times, in seconds, of
    Anglican and WebPPL, for $100\,000$ iterations of Markov Chain
    Monte Carlo (Lighweight Metropolis-Hastings), averaged over
    100 runs. Anglican outperforms WebPPL on both models.}
    \label{tbl:webppl}
\end{table}

In addition, \cite[pp. 32 -- 33]{P16} 
provides a comparison of Anglican to an older, interpreted
version of Anglican~\cite{WVM14}, and Probabilistic
Scheme~\cite{PW14} on a rather complicated inference task of
program synthesis. According to this comparison, Anglican is at
least 10 times faster than the older interpreted version, and
almost as fast as Probabilistic Scheme which uses inference
engine implemented in C. In addition, \cite[p. 171]{SGG15}
compares Anglican to a DSL for probabilistic programming in
Haskell and to Probabilistic C~\cite{PW14} on two simple
probabilistic programs. Anglican shows similar performance
to the Haskell implementation of Sequential Monte
Carlo~\cite{DFG+01} and scales better with the number of
particles. Given the simplicity of the programs used in the
evaluation, this comparison is a confirmation that the flexible
checkpoint and trampoline based interaction between an Anglican
program and the inference engine does not introduce any noticeable
overhead compared to more rigid designs.
